Air pollution and agricultural productivity in a developing country^[I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my

### 2023-04-11 draft/draft.Rmd
vastly updated draft

---
title: "Blowing in the wind: Pollution and agricultural productivity in India^[I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent Pollution is a growing problem in many developing countries. Using the rollout of coal plants to identify high-pollution locations in India and random variation induced by changes in wind direction, I show that pollution has large, negative impacts on agricultural productivity. Results indicate that some of this impact is driven by changes in labor allocation patterns in response to increased exposure to pollution. I also show that pollution from coal plants, specifically, has a larger effect on agricultural productivity than other types of pollution.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

# Tables
plantresultstable <- readRDS("pollution_development/draft/tables/plantresultstable.rds")
pollutiontable <- readRDS("pollution_development/draft/tables/pollutiontable.rds")
yield1reducedform <- readRDS("pollution_development/draft/tables/yield1reducedform.rds")
yield2naive <- readRDS("pollution_development/draft/tables/yield2naive.rds")
yield3ivmain <- readRDS("pollution_development/draft/tables/yield3ivmain.rds")
yield3ivmain_lead <- readRDS("pollution_development/draft/tables/yield3ivmain_lead.rds")
yield4monthly <- readRDS("pollution_development/draft/tables/yield4monthly.rds")
yield5heterogeneity <- readRDS("pollution_development/draft/tables/yield5heterogeneity.rds")
yield6leads <- readRDS("pollution_development/draft/tables/yield6leads.rds")
yield7ivdiffindiff <- readRDS("pollution_development/draft/tables/yield7ivdiffindiff.rds")
labor1all <- readRDS("pollution_development/draft/tables/labor1all.rds")
labor2rural <- readRDS("pollution_development/draft/tables/labor2rural.rds")
labor3month <- readRDS("pollution_development/draft/tables/labor3month.rds")
laborplot <- readRDS("pollution_development/draft/tables/laborplot.rds")
monthplot <- readRDS("pollution_development/draft/tables/monthplot.rds")
opplot <- readRDS("pollution_development/draft/tables/opplot.rds")
coef_vec <- readRDS("pollution_development/draft/tables/randomization_coef_vec.rds")
coefs_distfe <- readRDS("pollution_development/draft/tables/randomization_coefs_distfe.rds")
true_value <- readRDS("pollution_development/draft/tables/randomization_true_value.rds")

@@

Coal is one of the most common energy sources throughout the world. In 2021, energy generation from coal reached an all-time high,[^1] driven in part by rapid population growth and rising energy prices. India is no exception to these trends. From 1990 to 2010, India's population grew by more than 40 percent.[^2] This population growth led to a large increase in demand for power. While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. During those same two decades, power generation from all coal units of at least 30 megawatts in the country increased by almost 140 percent, from 42.4 gigawatts to more than 100 gigawatts.[^3] There are no doubt positive effects from this increase in power generation; for example, electricity may lead to economic and productivity growth [@dinkelman2011effects; @kline2014local; @rud2012electricity; @van2017long], though electrification alone may not be a sufficient condition [@lee2020does].

@@

These improvements do not come without downsides, however, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal [@iea2022]. Emissions from the production and burning of coal are also directly harmful to health and the environment, including pollutants like sulfur dioxide, nitrogen oxides, and mercury, to name but a few.[^4] While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.[^5]

@@

I study the effects of emissions from coal power plants on agricultural productivity in India. Previous work has suggested that there may be harmful direct effects of air pollution [@heck1982assessment; @marshall1997hidden] and water pollution [@reddy2006impact] on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana [@aragon2016polluting]. These effects can work through the well known effects on human health and productivity, but also directly through land productivity. Black carbon in soot, for example, can directly affect solar radiation reaching plants on the surface [@ramanathan2008global; @burney2014recent]. Relative to previous work in economics, this paper tackles the broader case of coal plants, which are a leading source of power throughout the world. I show that air pollution from these plants can lead to lower agricultural productivity and even decreased economic growth.

Identifying the causal effects of pollution is difficult. For starters, the construction of coal plants is likely to be endogenous, though the relationship between construction and economic conditions is not a priori clear. On the one hand, governments may locate new plants in fast-growing areas, while on the other hand, they may decide the exact location of plants based on the political power of local citizens [@kopas2020environmental]. In the case of the data used here, coal plants are constructed in areas with higher population and higher literacy (Table \@ref(tab:plantresultstable)). In terms of agriculture, higher productivity predicts more coal plants in a given year as well as the construction of a coal plant over the next decade, indicating possible bias when estimating OLS regressions of agricultural productivity on pollution levels.

To identify the effects of air pollution on agricultural productivity, I instead combine the roll out of coal plants in India with plausibly exogenous changes in wind direction -- in a similar spirit to @deryugina2019mortality -- to measure the effects of pollution from coal plants on key development outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 2000s to the 2010s, with differences across outcomes depending on data availability.

I use the location of coal plants to identify high-pollution areas and calculate the direction from each coal plant to all villages within 30km. I include all areas, regardless of whether the coal plant has opened yet, to allow me to disaggregate the effects of pollution from coal plants with other kinds of pollution. Using daily wind data and these spatial variables, I create an identifier separately for each of the one-hundred-thousand-plus villages for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. Consistent with expectations, wind direction is a strong predictor of particulate matter at the monthly level. More days being located downwind from a coal power plant leads to higher average concentrations of particulate matter 2.5.

Using newly available estimates of India-wide agricultural (land) productivity, results show that the total number of days of being downwind from a coal power plant has a negative effect on overall agricultural productivity. Just a single additional day of being downwind from a coal power plant leads to a decrease in predicted yields of 0.6 percent. Combined with the within-village absolute deviation of wind exposure, results indicate that median changes in wind direction are responsible for fluctuations in agricultural productivity of around 1.5 percent.

Using wind direction as an instrument for particulate matter, an increase of PM 2.5 by one percent leads to an even larger decrease in agricultural productivity, at around 0.73 percent. Given within-village changes across years, pollution could be responsible for swings of 7-to-14 percent in agricultural productivity across years. This IV strategy is similar to that in @deryugina2019mortality, though they use general "high-pollution" wind directions while I concentrate only on wind directions from the location of a coal power plant. The IV estimates indicate that the effect of pollution is more negative than naïve OLS estimates indicate, which is consistent with the evidence on the construction of coal plants; higher levels of pollution are correlated with higher levels of agricultural productivity and economic growth, a relationship that the IV strategy adjusts downwards.

These effects are apparently larger in villages that experience higher median exposure to pollution and in villages with higher initial yields. I also show that rainfall shocks are compounding; higher exposure along with less rainfall increases the negative effect of either shock individually. As weather shocks become more common across the globe, the rise of pollution in India may be particularly problematic for farmers.

To explore the mechanisms driving this effect, I also use data on labor allocation from the National Sample Survey (NSS). While there do not appear to be large changes in labor allocation patterns when looking at the season as a whole, there is a clear pattern when we break down exposure to pollution by month: higher levels of pollution at the beginning of the monsoon season (June and July) lead to lower levels of farm labor allocation, but no changes in non-farm labor allocation. This matches up well with a similar exercise with agricultural yields: pollution in June is the only month in which there is a negative effect of pollution on agricultural yield. In tandem, these two results point to effects on labor as one mechanism driving the total effects, though I cannot rule out direct effects of pollution on land productivity.

The final set of results tries to disentangle the effects of pollution from the effects of coal pollution, specifically. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.15 with respect to non-coal pollution. The opening of a coal plant, however, triples the elasticity, to more than -0.45, indicating that coal plant emissions are particularly harmful to agricultural productivity. This is consistent with coal emissions being quite different from other forms of pollution; for example, coal emissions contain much more mercury than other forms of pollution.

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. The papers most related to this one are @aragon2016polluting, who show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas, and @burney2014recent, who model the effects of climate change and pollutants on agricultural productivity in India. This paper differs in two important ways. First, I explore effects of pollution from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants.

@sanders2022adaptation

@@

# Data and methods

@@

In this section, I describe the different data sources and the identification strategy. I also spend time validating the use of wind direction as a proxy for particulate matter as well as looking at the correlates of construction of coal power plants.

## Data

\label{data2}

@@

## Identification

@@

So what are the characteristics of areas near the construction of new coal plants? Table \@ref(tab:plantresultstable) presents four separate regressions. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

# Table
colnames(plantresultstable) <- c("1991", "2001", "2001", "2011")
rownames(plantresultstable) <- c("Population (log)", "", "Area (log)", "", "Literacy (prop)", "", "Ag. productivity", "", 
                                "Sub-sample", "Observations")
kable(
      plantresultstable, caption = "Local characteristics and the construction of coal plants",
      align = "cccc", booktabs = TRUE, linesep = "", escape = FALSE
      ) %>%
  footnote(
           "Note: Robust standard errors are in parentheses. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01",
           general_title = "",
           threeparttable = TRUE,
           footnote_as_chunk = TRUE,
           escape = FALSE
           ) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(c(2:5),width = "2cm") %>%
  row_spec(9, hline_after = TRUE) %>%
  add_header_above(c(" ", "1991 census" = 2, "2001 census" = 2)) %>%
  kable_classic_2()

@@

The last two columns are particularly relevant to the current results. They present similar results to those in the first two columns, but use the 2001 census data and a dummy for the presence of a coal plant in 2001 and 2011 in columns three and four, respectively. Importantly, I include agricultural productivity in the monsoon season of 2002 -- the earliest year available -- as an additional predictor. In 2001, agricultural productivity is a strong predictor for the presence of a coal plant. Moreover, it is also a significant predictor of the construction of a coal plant between 2001 and 2011 (column four). Given that the coefficient is positive, this indicates that coal plants are more likely to be built in areas with higher agricultural productivity. In other words, any selection in the overall location of coal plants would bias against finding negative effects of pollution on agricultural productivity. However, note that this is not the same as looking at wind direction, though comparing the OLS and IV results below leads to the same conclusion regarding selection.

@@

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \@ref(tab:pollutiontable) of the appendix. The outcome variable is particulate matter -- specifically, $\mathrm{PM_{2.5}}$, which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from @hammer2020global. $\mathrm{PM_{2.5}}$ is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.[^12] This data is available at the monthly level, so in Table \@ref(tab:pollutiontable), I aggregate wind exposure to the monthly level, as well.

@@

In addition to serving as an example, Table \@ref(tab:pollutiontable) also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated $\mathrm{PM_{2.5}}$ is substantially higher.

To put these numbers into perspective, the World Health Organization's updated guidelines are that average annual exposure should not exceed 15 $\mathrm{\mu g/m^3}$. For the month, an increase at the midpoint of the estimated range is equal to an increase of 0.2 percent of the maximum recommended mean concentration from the WHO. This is just a single day of wind in the direction of a village. Since this includes all villages located within 30km of a coal plant, this is evidence that wind direction can have serious repercussions on the health of hundreds of millions of people in India.

### Identifying effects of pollution from coal plants

@@

This section presents the main results of this paper. I present two main sets of results: one set looking at agricultural productivity and another looking at labor allocation.

## Agricultural productivity

@@

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \@ref(tab:yieldtablepm) presents the same regressions as Table \@ref(tab:yieldtable), with a key difference: instead of wind exposure, it uses log average concentration of $\mathrm{PM_{2.5}}$ -- from @hammer2020global -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.03-to-0.04 percent decrease in agricultural productivity.

@@

The second-stage results first indicate that $\mathrm{PM_{2.5}}$ is a very strong predictor of agricultural productivity. Using the results in column three, a one percent increase in particulate matter leads to a 0.73 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of 7-to-14 percent in agricultural productivity due to pollution exposure are common.

@@

@@

The first two columns split the sample based on median exposure. In other words, for each village, I calculate the median exposure (based on wind) and then use this single value for each village to split the sample into villages above and below median exposure. The coefficient in the first column is 2.3 times larger than the coefficient in the second column, indicating much larger effects of pollution in places where it is more common. One possible explanation is that exposure is cumulative, leading to larger and larger effects. I return to this in column five.

Columns three and four split the sample based on median yield at the very beginning of the sample (the first year available, 2002-2003). Column three presents results for villages above the median while column four presents results for villages below the median. The largest effects are apparently concentrated in the upper half of the distribution, meaning the most productive areas are also most susceptible to the effects of pollution.

@@

>>>>

---
title: "Air pollution and agricultural productivity in a developing country^[I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent I document large negative externalities in the agricultural sector of India due to air pollution. Using random variation in pollution exposure induced by changes in wind direction, I show that higher levels of particulate matter lead to decreased agricultural productivity. Based on within-village average deviations in pollution exposure, year-to-year changes in productivity can be above 10 percent. Combining wind direction with the rollout of coal plants, I show that pollution from coal plants, specifically, has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the effects of pollution may hit the poorest particularly hard.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

# Tables
plantresultstable <- readRDS("pollution_development/draft/tables/plantresultstable.rds")
pollutiontable <- readRDS("pollution_development/draft/tables/pollutiontable.rds")
yield1reducedform <- readRDS("pollution_development/draft/tables/yield1reducedform.rds")
yield2naive <- readRDS("pollution_development/draft/tables/yield2naive.rds")
yield3ivmain <- readRDS("pollution_development/draft/tables/yield3ivmain.rds")
yield3ivmain_lead <- readRDS("pollution_development/draft/tables/yield3ivmain_lead.rds")
yield4monthly <- readRDS("pollution_development/draft/tables/yield4monthly.rds")
yield5heterogeneity <- readRDS("pollution_development/draft/tables/yield5heterogeneity.rds")
yield6leads <- readRDS("pollution_development/draft/tables/yield6leads.rds")
yield7ivdiffindiff <- readRDS("pollution_development/draft/tables/yield7ivdiffindiff.rds")
yield8cumulative <- readRDS("pollution_development/draft/tables/yield8cumulative.rds")
labor1all <- readRDS("pollution_development/draft/tables/labor1all.rds")
labor2rural <- readRDS("pollution_development/draft/tables/labor2rural.rds")
labor3month <- readRDS("pollution_development/draft/tables/labor3month.rds")
laborplot <- readRDS("pollution_development/draft/tables/laborplot.rds")
monthplot <- readRDS("pollution_development/draft/tables/monthplot.rds")
opplot <- readRDS("pollution_development/draft/tables/opplot.rds")
coef_vec <- readRDS("pollution_development/draft/tables/randomization_coef_vec.rds")
coefs_distfe <- readRDS("pollution_development/draft/tables/randomization_coefs_distfe.rds")
true_value <- readRDS("pollution_development/draft/tables/randomization_true_value.rds")
coef_current <- readRDS("pollution_development/draft/tables/coef_current.rds")
coef_lead <- readRDS("pollution_development/draft/tables/coef_lead.rds")

@@

While reliance on coal is declining in much of the world, it continues to be an important source of energy in many countries. In 2021, energy generation from coal reached an all-time high,[^1] driven in part by rapid population growth in developing countries, including India. From 1990 to 2010, India's population grew by more than 40 percent, leading to a large increase in demand for power.[^2] While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. During those same two decades, power generation from all coal units of at least 30 megawatts in the country more than doubled, from 42.4 gigawatts to more than 100 gigawatts.[^3]

@@

There are no doubt positive effects from this increase in power generation; for example, electricity may lead to economic and productivity growth [@dinkelman2011effects; @kline2014local; @rud2012electricity; @van2017long], though electrification alone may not be a sufficient condition [@lee2020does]. These improvements do not come without downsides, however, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal [@iea2022]. Emissions from the production and burning of coal are also directly harmful to health and the environment, including pollutants like sulfur dioxide, nitrogen oxides, and mercury, to name but a few.[^4] While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.[^5]

@@

In addition to coal, pollution more generally is increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.^[https://edgar.jrc.ec.europa.eu] Of course, much of this increase is due to the development process itself; as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services [@herrendorf2014growth]. Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor [@dercon2009rural]. This also means agriculture is a key sector when it comes to poverty reduction [@christiaensen2011evolving]. Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

In this paper, I study the effects of pollution on agricultural productivity in India. Previous work has suggested that there may be harmful direct effects of air pollution [@heck1982assessment; @marshall1997hidden] and water pollution [@reddy2006impact] on agricultural productivity, including work showing that gold mines -- which release large amounts of pollutants into the surrounding environment -- lead to lower agricultural productivity in Ghana [@aragon2016polluting]. These effects can work through the well-known effects on human health and productivity, but also directly through land productivity. Black carbon in soot, for example, can directly affect solar radiation reaching plants on the surface [@ramanathan2008global; @burney2014recent]. Relative to previous work in economics, this paper tackles pollution more broadly, before turning to a specific source of pollution: coal plants. I show that air pollution leads to lower agricultural productivity in India and that pollution from coal plants is particularly harmful to agriculture.

Identifying the causal effects of pollution is difficult. For starters, the construction of coal plants is likely to be endogenous, though the relationship between construction and economic conditions is not a priori clear. On the one hand, governments may locate new plants in fast-growing areas, while on the other hand, they may decide the exact location of plants based on the political power of local citizens [@kopas2020environmental]. In the case of the data I use here, coal plants are constructed in areas with higher population and higher literacy. In terms of agriculture, higher productivity predicts more coal plants in a given year as well as the construction of a coal plant over the next decade, indicating possible bias when estimating OLS regressions of agricultural productivity on pollution levels.

To identify the effects of air pollution on agricultural productivity, I identify areas of high pollution emission using the location of coal plants. While coal plants may add emissions, I show that they are located in areas with relatively higher levels of pollution, in general. Using these locations as point sources of pollution, I use plausibly exogenous changes in wind direction -- in a similar spirit to @deryugina2019mortality -- to measure the effects of pollution on key agricultural outcomes. I first plot the location of coal plants in India by year. Using a village-level shapefile, I identify all villages that are located within 30km of a coal plant at any time during the sample period, which comprises the 2000s to the 2010s, with differences across outcomes depending on data availability. I then calculate the direction from each coal plant to all villages within 30km. I include all areas, regardless of whether the coal plant has opened yet, to allow me to disaggregate the effects of pollution from coal plants with other kinds of pollution later in the paper. Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5).^[PM 2.5 is defined as particulate matter with a diameter of less than 2.5 micrometers.]

Using wind direction as an instrument for particulate matter, an increase of PM 2.5 by one percent leads to a large decrease in agricultural productivity, at around 0.73 percent. Given within-village changes in exposure, pollution could be responsible for swings of 7-to-14 percent in agricultural productivity across years. The IV estimates indicate that the effect of pollution is more negative than naïve OLS estimates, which is consistent with the evidence on the construction of coal plants; higher levels of pollution are correlated with higher levels of agricultural productivity, a relationship that the IV strategy adjusts downwards. Robustness checks confirm these main results. Randomization inference shows that the estimates are robust to alternative measures of uncertainty, while leads of pollution do not affect current agricultural productivity.

Heterogeneity analysis shows that the negative effects of pollution are larger in villages that experience higher median exposure to pollution and in villages with higher initial yields. Finally, given how many poor farmers rely on rainfall and how weather shocks increase across the globe, I also look at possible multiplicative effects of shocks by interacting the pollution variables with a rainfall variable. Higher exposure along with less rainfall increases the negative effect of either shock individually, meaning these shocks are compounding. Given concerns regarding agricultural productivity and climate change, the rise of pollution in India may be particularly problematic for poor farmers.

I find that these effects are mostly transmitted by reductions in farm labor during narrow windows of the growing season, using labor allocation data from the National Sample Survey (NSS). While there do not appear to be large changes in labor allocation patterns when looking at the season as a whole, there is a clear pattern when we break down exposure to pollution by month: higher levels of pollution at the beginning of the monsoon season (June and July) lead to lower levels of farm labor allocation, but no changes in non-farm labor allocation. This matches up well with a similar exercise with agricultural yields: pollution in June is the only month in which there is a negative effect of pollution on agricultural yield. In tandem, these two results point to effects on labor as one mechanism driving the total effects, though I cannot rule out direct effects of pollution on land productivity with my data.

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.15 with respect to non-coal pollution. The opening of a coal plant, however, triples the elasticity, to more than -0.45, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; for example, coal emissions contain much more mercury than other forms of pollution.

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. Several papers are closely related to this one. @aragon2016polluting show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas. @burney2014recent model the effects of climate change and pollutants on agricultural productivity in India. Finally, @sanders2022adaptation actually find that coal emissions can be good for agriculture in the United States, due to the presence of sulfur, which can increase agricultural productivity. This paper differs in three important ways. First, I explore effects of pollution more generally as well as from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants. Finally, agriculture in India and the United States can be quite different, with the latter being much more mechanized. This means effects of pollution could differ substantially.

@@

# Data

@@

@@

# Methods

@@

So what are the characteristics of areas near the construction of new coal plants? Table \@ref(tab:plantresultstable) presents four separate regressions, using state-level fixed effects due to the spatial variation in coal plants. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

# Table
colnames(plantresultstable) <- c("1991", "2001", "2001", "2011")
rownames(plantresultstable) <- c("Population (log)", "", "Area (log)", "", "Literacy (prop)", "", "Ag. productivity", "", "Pollution", "(log PM 2.5)",
                                "Sub-sample", "Observations")
kable(
      plantresultstable, caption = "Local characteristics and the construction of coal plants",
      align = "cccc", booktabs = TRUE, linesep = "", escape = FALSE
      ) %>%
  footnote(
           "Note: State fixed effects are included in regressions. Standard errors are in parentheses and clustered at the state level. The outcome in the first column is whether the village is within 30km of a coal plant in 1991. The second column is whether a village in 1991 will have a coal plant in 2001, conditional on not having one in 1991. The last two columns are similarly defined, except using 2001 and 2011 as the years. * p<0.10 ** p<0.05 *** p<0.01",
           general_title = "",
           threeparttable = TRUE,
           footnote_as_chunk = TRUE,
           escape = FALSE
           ) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(c(2:5),width = "2cm") %>%
  row_spec(11, hline_after = TRUE) %>%
  add_header_above(c(" ", "1991 census" = 2, "2001 census" = 2)) %>%
  kable_classic_2()

@@

The last two columns are particularly relevant to the current results. They present similar results to those in the first two columns, but use the 2001 census data and a dummy for the presence of a coal plant in 2001 and 2011 in columns three and four, respectively. Importantly, I include agricultural productivity in the monsoon season of 2002 -- the earliest year available -- as well as average pollution during the monsoon season for 1998 through 2000 as additional predictors. In 2001, agricultural productivity is a strong predictor for the presence of a coal plant. Moreover, it is also a significant predictor of the construction of a coal plant between 2001 and 2011 (column four). Given that the coefficient is positive, this indicates that coal plants are more likely to be built in areas with higher agricultural productivity. In other words, any selection in the overall location of coal plants would bias against finding negative effects of pollution on agricultural productivity. However, note that this is not the same as looking at wind direction, though comparing the OLS and IV results below leads to the same conclusion regarding selection. Pollution, on the other hand, is not significant but is positive in both columns (and is also logged, just like productivity).

@@

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \@ref(tab:pollutiontable) of the appendix. The outcome variable is particulate matter -- specifically, $\mathrm{PM 2.5}$, which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from @hammer2020global. $\mathrm{PM 2.5}$ is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.[^12] This data is available at the monthly level, so in Table \@ref(tab:pollutiontable), I aggregate wind exposure to the monthly level, as well.

@@

In addition to serving as an example, Table \@ref(tab:pollutiontable) also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated $\mathrm{PM 2.5}$ is substantially higher.

To put context behind pollution numbers, the World Health Organization's updated guidelines are that average annual exposure should not exceed 15 $\mathrm{\mu g/m^3}$. For the month, an increase at the midpoint of the estimated range is equal to an increase of 0.2 percent of the maximum recommended mean concentration from the WHO. This is just a single day of wind in the direction of a village. Since this includes all villages located within 30km of a coal plant, this is evidence that wind direction can have serious repercussions on the health of hundreds of millions of people in India.

## Identifying effects of pollution from coal plants

@@

@@

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \@ref(tab:yieldtablepm) presents the same regressions as Table \@ref(tab:yieldtable), with a key difference: instead of wind exposure, it uses log average concentration of $\mathrm{PM 2.5}$ -- from @hammer2020global -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.03-to-0.04 percent decrease in agricultural productivity.

@@

The second-stage results first indicate that $\mathrm{PM 2.5}$ is a very strong predictor of agricultural productivity. Using the results in column three, a one percent increase in particulate matter leads to a 0.73 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of 7-to-14 percent in agricultural productivity due to pollution exposure are common.

@@

In this section, I report the results of two separate robustness checks for the main results presented above. In the previous results, standard errors are clustered at the level of the village, which is also the level at which treatment (exposure to pollution) is assigned as well as the level at which agricultural productivity is extracted. Past work has shown that standard errors might be underestimated in such contexts [@bertrand2004much]. This is a particular concern in the present case, since agricultural productivity is extracted at the village level but estimated using less coarse data that has high spatial correlation [@gangopadhyay2022new].

To help remedy concerns, I perform randomization inference (e.g. @barrios2012clustering) by randomly assigning observed treatment across villages and repeating this process 1,000 times. I do this in two separate ways. First, I take all villages in a given year and randomly assign, with replacement, treatment intensity within the year. Second, I do this separately by selecting from within district-years only, to help take into account district-year correlation in outcomes and as a way to mitigate possible concerns that traditional clustered standard errors may actually be much too conservative [@abadie2023should]. Since this is an instrumental variables procedure, I randomly select the wind and pollution variable _together_.

I present these results in Figure \@ref(fig:randomplot). The purple density is the resulting distribution of estimated treatment effects when selecting from all villages within the same year. The density of treatment effects is relatively tight around 0, with the left tail only getting slightly beyond -0.05, compared to the most conservative actual estimated treatment effect of -0.73.

The within-district correlation in outcomes and treatment is apparent when instead randomly selecting from only villages within the same district in the same year. The distribution is still approximately normal but has a mode of almost -0.4. Nonetheless, the actual treatment effect is larger in magnitude than every single one of the 1,000 bootstrapped replications. If we were to use the standard deviation of the resulting distribution as a standard error, we would get a confidence interval around the main effect of approximately `r paste0(reportDec(-0.73-1.96*sd(coefs_distfe)))` to `r paste0(reportDec(-0.73+1.96*sd(coefs_distfe)))`. The results are apparently robust to alternative specifications of uncertainty

The second robustness check includes leads of pollution and wind direction. I present these results in Table \@ref(tab:yieldtableleads). Interestingly, there is actually a positive coefficient for the lead in the second column. However, if we perform randomization inference, it appears that the confidence intervals might be slightly underestimated; the positive effect size is in the 11th percentile of the distribution in Figure \@ref(fig:randomplotleads), which for a two-tailed test would not even have a p-value of less than 0.2. Overall, there is no indication that leads predict any changes in current agricultural productivity.

@@

The first two columns split the sample based on median exposure. In other words, for each village, I calculate the median exposure (based on wind) and then use this single value for each village to split the sample into villages above and below median exposure. The coefficient in the first column is 2.3 times larger than the coefficient in the second column, indicating much larger effects of pollution in places where it is more common. One possible explanation is that exposure is cumulative, leading to larger and larger effects. I return to the effects of cumulative shocks in column five.

Columns three and four split the sample based on median yield at the very beginning of the sample (the first year available, 2002-2003). Column three presents results for villages above the median while column four presents results for villages below the median. The largest effects are apparently concentrated in the upper half of the distribution, meaning the most productive areas are also most susceptible to the effects of pollution. This could have important implications for the overall effects of the placement of any point sources of pollution.

@@

```{r randomplotleads, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 8, fig.cap = "Randomization inference for leads", fig.footer = list("Note: The figure presents the distribution of coefficients by randomly assigning a single wind value to each village, sampled with replacement across all villages in a given district in the same year. \"Current\" refers to the coefficient on current pollution, while \"lead\" refers to the coefficient on next year's pollution in the same table.")}
